Efficient simulation of individual-based population models

### 2024-03-26 article.qmd
update

@@

Stochastic IBMs allow the modeling in continuous time of populations dynamics structured by age and/or characteristics. In the field of mathematical biology and ecology, a large community has used this formalism for the study of the evolution of structured populations (see e.g. [@FerTra09; @collet2013rigorous; @BanMel15; @costa2016stochastic; @billiard2016effect; @lavallee2019stochastic; @meleard2019birth; @calvez2020horizontal]), after the pioneer works [@FouMel04; @champagnat2006unifying; @tran_2008].

IBMs are also useful in demography and actuarial sciences, for the modeling of human populations dynamics (see e.g. [@Ben10; @Bou16; @karoui2021simulating]). They allow the modeling of heterogeneous and complex population dynamics, which can be used to compute demographic indicators or simulate the evolution of insurance portfolios in order to study the basis risk, compute cash flows for annuity products or pension schemes, or for a fine assessment of mortality models [@barrieu2012understanding]. There are other domains in which stochastic IBMs can be used, for example in epidemiology with stochastic compartmental models, neurosciences, cyber risk, or Agent-Based Models (ABMs) in economy and social sciences, which can be seen as IBMs.
Many mathematical results have been obtained in the literature cited above, for quantifying the limit behaviors of IBMs in long time or in large population. In particular, pathwise representations of IBMs have been introduced in [@FouMel04] (and extended to age-structured populations in [@tran_2008]), as measure-valued pure jumps Markov processes, solutions of SDEs driven by Poisson measures. These pathwise representations are based on the *thinning* and projection of Poisson random measures defined on extended spaces. However, the simulation of large and interacting populations is often referred as computationally expensive.

@@

In the remainder of this section,  we define rigorously the class of IBMs that can be simulated in `IBMPopSim`, along with the assumptions that are required in order for the population to be simulatable. The representation of age-structured IBMs based on measure-valued processes, as introduced in [@tran_2008], is generalized to a wider class of abstract population dynamics. The modeling differs slightly here, since individuals are _kept in the population_ after their death (or exit), by including the death/exit date as an individual trait.

@@

3. `IBMPopSim` can simulate IBMs that include intensities expressed as a sum of Poisson intensities and individual intensities of the form $\Lambda^e(Z_t) =\mu^e_t + \sum_{k=1}^{N_t} \lambda^e(I_k, Z_t)$. Other examples are provided in @sec-insurance-portfolio and @sec-example-interaction.
Finally, the global intensity at which an event can occur in the population is defined by: 
$$
    \Lambda_t(Z_t) = \sum_{e\in \mathcal{P}} \mu^e(t) + \sum_{e \in \mathcal E} \Big(\sum_{k=1}^{N_t} \lambda^e(t, I_k)\Big) + \sum_{e \in \mathcal E_W} \Big(\sum_{k=1}^{N_t}\sum_{j=1}^{N_t} W^e(t, I_k, I_j)\Big).
$$ {#eq-global-ev-intensity}
An important point is that for events $e \in \mathcal E$ without interactions, the global event intensity $\Lambda^e_t(Z_t) = \sum_{k=1}^{N_t} \lambda^e(t, I_k)$ is _of order_ $N_t^a$ defined in @eq-Nta (number of alive individuals at time $t$). On the other hand, for events $e \in \mathcal{E}_W$ with interactions, $\Lambda^e_t(Z_t) = \sum_{k=1}^{N_t}\sum_{j=1}^{N_t} W^e(t, I_k, I_j)$ is of order $(N_t^a)^2$. 
Informally, this means that when the population size increases, events with interaction are more costly to simulate. Furthermore, the numerous computations of the interaction kernel $W^e$ can also be quite costly. The randomized Algorithm [\[algo::rzndomized\]](#algo::rzndomized){reference-type="ref" reference="algo::rzndomized"}, detailed in Section [2.3](#sec::simulation_algo_randomized){reference-type="ref" reference="sec::simulation_algo_randomized"}, allows us to overcome these limitations.

@@

We now present the main algorithm for simulating the evolution of an IBM over $[0,T]$.The algorithm implemented in `IBMPopSim` allows the exact simulation of @eq-sde-eq-Z, based on an acceptance/reject algorithm for simulating random times called *thinning*. The exact simulation of event times with this acceptance/reject procedure is closely related to the simulations of inhomogeneous Poisson processes by the so-called thinning algorithm, often attributed to [@LewShe79]. 
The simulation methods for inhomogeneous Poisson processes can be adapted to IBMs, and we introduce in this section a general algorithm extending those by [@FouMel04] (see also [@FerTra09], [@Ben10]).

@@

![test](img/plot_thinning.tex){#fig-thinning}

@@

Let $e \in \cal E_W$ be an event with interactions. In order to evaluate the individual intensity $\lambda^e_t (I,Z_t) = \sum_{j=1}^{N_t} W^e(t, I,I_j)$ one must compute $W^e(t, I_l, I_j)$ for all individuals in the population. This step can be computationally costly, especially for large populations. 
One way to avoid this summation is to use randomization (see also [@FouMel04] in a model without age). The randomization consists in replacing the summation by an evaluation of the interaction function $W^e$ using an individual $J$ drawn uniformly from the population.

@@

The argument `interaction_type`, set by default at `random`, is the algorithm choice for simulating the model. When `interaction_type=full`, the simulation follows @alg-no-interaction, `interaction_type=random` it follows @alg-interaction.
In most cases, the `random` algorithm is much faster than the `full` algorithm, as we illustrate for instance in @sec-example-interaction , where we observe the gain of a factor of 40 between the two algorithms, on a set of standard parameters. This allows in particular to explore parameter sets that give larger population sizes, without reaching computation times that explode.

@@

@@

>>>>

```{r echo=FALSE}
library(ggplot2)
```

@@

Stochastic IBMs allow the modeling in continuous time of populations dynamics structured by age and/or characteristics. In the field of mathematical biology and ecology, a large community has used this formalism for the study of the evolution of structured populations (see e.g. [@FerTra09; @collet2013rigorous; @BanMel15; @costa2016stochastic; @billiard2016effect; @lavallee2019stochastic; @meleard2019birth; @calvez2020horizontal]), after the pioneer works [@FouMel04;@tran_2008;@meleard2009trait].

IBMs are also useful in demography and actuarial sciences, for the modeling of human populations dynamics (see e.g. [@Ben10; @Bou16; @karoui2021simulating]). They allow the modeling of heterogeneous and complex population dynamics, which can be used to compute demographic indicators or simulate the evolution of insurance portfolios in order to study the basis risk, compute cash flows for annuity products or pension schemes, or for a fine assessment of mortality models [@barrieu2012understanding]. There are other domains in which stochastic IBMs can be used, for example in epidemiology with stochastic compartmental models, neurosciences, cyber risk, or Agent-Based Models (ABMs) in economy and social sciences, which can be seen as IBMs.
Many mathematical results have been obtained in the literature cited above, for quantifying the limit behaviors of IBMs in long time or in large population. In particular, pathwise representations of IBMs have been introduced in [@FouMel04] (and extended to age-structured populations in [@tran_2008;@meleard2009trait]), as measure-valued pure jumps Markov processes, solutions of SDEs driven by Poisson measures. These pathwise representations are based on the *thinning* and projection of Poisson random measures defined on extended spaces. However, the simulation of large and interacting populations is often referred as computationally expensive.

@@

In the remainder of this section,  we define rigorously the class of IBMs that can be simulated in `IBMPopSim`, along with the assumptions that are required in order for the population to be simulatable. The representation of age-structured IBMs based on measure-valued processes, as introduced in @tran_2008, is generalized to a wider class of abstract population dynamics. The modeling differs slightly here, since individuals are _kept in the population_ after their death (or exit), by including the death/exit date as an individual trait.

@@

3. `IBMPopSim` can simulate IBMs that include intensities expressed as a sum of Poisson intensities and individual intensities of the form $\Lambda^e(Z_t) =\mu^e_t + \sum_{k=1}^{N_t} \lambda^e(I_k, Z_t)$. Other examples are provided in @sec-insurance-portfolio and @sec-example-interaction.
Finally, the global intensity at which an event can occur in the population is defined by: 
$$
    \Lambda_t(Z_t) = \sum_{e\in \mathcal{P}} \mu^e(t) + \sum_{e \in \mathcal E} \Big(\sum_{k=1}^{N_t} \lambda^e(t, I_k)\Big) + \sum_{e \in \mathcal E_W} \Big(\sum_{k=1}^{N_t}\sum_{j=1}^{N_t} W^e(t, I_k, I_j)\Big).
$$ {#eq-global-ev-intensity}
An important point is that for events $e \in \mathcal E$ without interactions, the global event intensity $\Lambda^e_t(Z_t) = \sum_{k=1}^{N_t} \lambda^e(t, I_k)$ is _of order_ $N_t^a$ defined in @eq-Nta (number of alive individuals at time $t$). On the other hand, for events $e \in \mathcal{E}_W$ with interactions, $\Lambda^e_t(Z_t) = \sum_{k=1}^{N_t}\sum_{j=1}^{N_t} W^e(t, I_k, I_j)$ is of order $(N_t^a)^2$. 
Informally, this means that when the population size increases, events with interaction are more costly to simulate. Furthermore, the numerous computations of the interaction kernel $W^e$ can also be quite costly. The randomized Algorithm @alg-randomized, detailed in Section [2.3](#sec::simulation_algo_randomized){reference-type="ref" reference="sec::simulation_algo_randomized"}, allows us to overcome these limitations.

@@

We now present the main algorithm for simulating the evolution of an IBM over $[0,T]$.The algorithm implemented in `IBMPopSim` allows the exact simulation of @eq-sde-eq-Z, based on an acceptance/reject algorithm for simulating random times called *thinning*. The exact simulation of event times with this acceptance/reject procedure is closely related to the simulations of inhomogeneous Poisson processes by the so-called thinning algorithm, often attributed to @LewShe79. 
The simulation methods for inhomogeneous Poisson processes can be adapted to IBMs, and we introduce in this section a general algorithm extending those by @FouMel04 (see also @FerTra09, @Ben10).

@@

![Realization of a Marked Poisson measure $\bar Q$ on $[0,T]$ with mean measure $\bar \mu(\mathrm{d} t, \mathrm{d} \theta) = \mathrm{d} t \mathbf{1}_{[0, \bar \Lambda]}(\theta) \mathrm{d} \theta$ (red crosses), and realization of the restriction $\bar Q^\Delta$ where $\Delta = \{ (t,\theta)\in [0,T]\times[0, \bar \Lambda], \theta \leq \Lambda(t) \}$ (blue circles). The projection of $\bar Q^\Delta$ on first component is an inhomogeneous Poisson process on $[0,T]$ of intensity $(\Lambda(t))$ and jump times $(T_k)_{k \ge 1}$.](img/plot_thinning.tex){#fig-thinning}

@@

Let $e \in \cal E_W$ be an event with interactions. In order to evaluate the individual intensity $\lambda^e_t (I,Z_t) = \sum_{j=1}^{N_t} W^e(t, I,I_j)$ one must compute $W^e(t, I_l, I_j)$ for all individuals in the population. This step can be computationally costly, especially for large populations. 
One way to avoid this summation is to use randomization (see also @FouMel04 in a model without age). The randomization consists in replacing the summation by an evaluation of the interaction function $W^e$ using an individual $J$ drawn uniformly from the population.

@@

The argument `interaction_type`, set by default at `random`, is the algorithm choice for simulating the model. When `interaction_type=full`, the simulation follows @alg-interaction, `interaction_type=random` it follows @alg-randomized.
In most cases, the `random` algorithm is much faster than the `full` algorithm, as we illustrate for instance in @sec-example-interaction , where we observe the gain of a factor of 40 between the two algorithms, on a set of standard parameters. This allows in particular to explore parameter sets that give larger population sizes, without reaching computation times that explode.

@@

This section provides an example of how to use the `IBMPopSim` package to simulate an age-structured population with interactions, based on the model proposed in Example 1 of @FerTra09 (see also @meleard2009trait).

In this model, individuals are characterized by their body size at birth
$x_0 \in [0,4]$ and by their physical age $a \in [0,2]$. The body size
of an individual $I=(\tau^b,\infty, x_0)$ at time $t$ is a linear
function of its age $a(I,t) = t-\tau^b$: $$x(t)= x_0 + ga(I,t),$$ where
$g$ is a constant growth rate assumed to be identical for all
individuals.

**Birth events** The birth intensity of each individual
$I=(\tau^b, \infty, x_0)$ depends on a parameter $\alpha > 0$ and on its
initial size, as given by the equation
$$
\lambda^b(t,I) = \alpha (4 - x_0) \leq \bar \lambda^b = 4\alpha.
$${#eq-interaction-birth-intensity}
Thus,
smaller individuals have a higher birth intensity. When a birth occurs,
the new individual inherit the same birth size $x_0$ as its parent with
high probability $1-p$, or a mutation can occur with probability $p$,
resulting in a birth size given by
$$
    x_0' = \min(\max(0, x_0 + G), 4),
$${#eq-interaction-birth-kernel}
where $G$ is a Gaussian random
variable with mean 0 and variance $\sigma^2$.

**Death events** Due to competition between individuals, the death
intensity of an individual depends on the size of other individuals in
the population. Bigger individuals have a better chance of survival. If
an individual $I= (\tau^b, \infty, x_0)$ of size $x(t)= x_0 +ga(I,t)$
encounters an individual $J= (\tau^{b}_J, \infty, x_0')$ of size
$x'(t) = x_0'+ ga(J,t)$, then it can die with the intensity
$$W(t, I,J) = U(x(t),x'(t)),$$ where the interaction function $U$ is
defined by 
$$
    U(x,y) = \beta \left(1- \frac{1}{1+ c\exp(-4(x-y))}\right) \leq \bar W = \beta.
$${#eq-interaction-death-intensity}
The death intensity of an individual $I$ at time $t$ and in a population
$Z$ is the result of interactions with all individuals in the
population, including itself, and is given by
$$\lambda^d_t(I,Z) = \sum_{J = (\tau^b,\infty, x_0') \in Z}  W (x_0 + g a(I,t), x_0' + g a(J,t)),$$

## Population

We use an initial population of 900 living individuals, all of whom have
the same size and ages uniformly distributed between 0 and 2 years.

```{r}
N <- 900 
x0 <- 1.06
agemin <- 0.
agemax <- 2.
```

```{r}
pop_df <- data.frame(
  "birth" = -runif(N, agemin, agemax), # Uniform age in [0,2]
  "death" = as.double(NA), # All individuals are alive
  "birth_size" = x0) # All individuals have the same initial birth size x0
pop_init <- population(pop_df)
```

## Events

### Birth events

The parameters involved in a birth event are the probability of mutation
$p$, the variance of the Gaussian random variable and the coefficient
$\alpha$ of the intensity.
```{r}
params_birth <- list("p" = 0.03, "sigma" = sqrt(0.01), "alpha" = 1)
```

The birth
intensity @eq-interaction-birth-intensity is of class `individual`.
Hence, the event is created by calling the `mk_event_individual`
function. The size of the new individual is given in the kernel
following @eq-interaction-birth-kernel.

```{r}
birth_event <- mk_event_individual( 
  type = "birth",
  intensity_code = "result = alpha*(4 - I.birth_size);",
  kernel_code = "if (CUnif() < p)
                   newI.birth_size = min(max(0.,CNorm(I.birth_size,sigma)),4.);
                 else
                   newI.birth_size = I.birth_size;")
```

### Death events

The death
intensity @eq-interaction-death-intensity is of class `interaction`.
Hence, the event is created by calling the `mk_event_interaction`
function. The parameters used for this event are the growth rate $g$,
the amplitude of the interaction function $\beta$, and the strength of
competition $c$.

```{r}
params_death <- list("g" = 1, "beta" = 2./300., "c" = 1.2)
death_event <- mk_event_interaction(
  type = "death",
  interaction_code = "double x_I = I.birth_size + g * age(I,t);
                      double x_J = J.birth_size + g * age(J,t);
                      result = beta*(1.-1./(1.+c*exp(-4.*(x_I-x_J))));")
```

## Model creation and simulation

The model is created using the `mk_model` function.

The simulation of one scenario can then be launched with the call of the
`popsim` function, after computing the events bounds
$\bar \lambda^b=4 \alpha$ and $\bar W= \beta$.

```{r}
model <- mk_model(
    characteristics = get_characteristics(pop_init),
    events = list(birth_event, death_event),
    parameters = c(params_birth, params_death))
```

The simulation of one scenario can then be launched with the call of the
`popsim` function, after computing the events bounds
$\bar \lambda^b=4 \alpha$ and $\bar W= \beta$.

```{r}
sim_out <- popsim(model = model,
    initial_population = pop_init,
    events_bounds = c("birth" = 4 * params_birth$alpha, 
                      "death" = params_death$beta),
    parameters = c(params_birth, params_death),
    age_max = 2,
    time = 500)
```
Based on the results of a simulation, we can reproduce the numerical
results of @FerTra09. In @fig-birth-size, we draw a line for each individual in the population to represent their birth size during their lifetime.

```{r, echo=FALSE, fig.width=10}
#| label: fig-birth-size
#| fig-cap: "Evolution of birth size"
#| fig-pos: 'h'

pop_out <- sim_out$population

ggplot(pop_out) + geom_segment(
  aes(x=birth, xend=death, y=birth_size, yend=birth_size),
  na.rm=TRUE, colour="blue", alpha=0.2) +
  xlab("Time") +
  ylab("Birth size")
```

In this example, the randomized
@alg-randomized allows for much faster computation
times than the model implemented below with
@alg-interaction ("full" algorithm):

```{r}
death_event_full <- mk_event_interaction(type = "death",
                             interaction_type= "full",
                             interaction_code = "double x_I = I.birth_size + g * age(I,t);
                                                 double x_J = J.birth_size + g * age(J,t);
                                result = beta * ( 1.- 1./(1. + c * exp(-4. * (x_I-x_J))));"
)

model_full <- mk_model(characteristics = get_characteristics(pop_init),
                       events = list(birth_event, death_event_full),
                       parameters = c(params_birth, params_death))

sim_out_full <- popsim(model = model_full,
                       initial_population = pop_init,
                       events_bounds =c("birth" = 4 * params_birth$alpha,
                                        "death" = params_death$beta),
                       parameters = c(params_birth, params_death),
                       age_max = 2,
                       time = 500)

```

```{r echo=FALSE}

print(paste("The full algorithm is " ,sim_out_full$logs["duration_ns"]/sim_out$logs["duration_ns"], " times slower than the randomized version", sep=""))
```

In @fig-fullvsrandom, the two algorithms are compared for different population sizes.  
We progressively decrease the value of the mortality rate parameter $\beta$ and increase the birth
rate parameter $\alpha$. Starting with the values provided
in @FerTra09, $\alpha=1$ and $\beta=2/300$, resulting in a stationary
population size of approximately $N=360$ individuals for a sample of 50
simulations, we can easily increase the stationary population size to
approximately $N=2600$ individuals with $\alpha=2$ and $\beta=1/300$
.^[The choices
    $(\alpha, \beta) \in \{(1,2/300),(1, 1/300), (1.5, 1/300), (2, 1/300)\}$
    lead to the stationary population sizes
    $N \in \{360, 900, 1800, 2600\}$. For each set of parameters, we
    generated a new initial population, which was used for a benchmark
    of 50 simulations with both randomized and full algorithm. The
    simulations run on a Intel Core i7-8550U CPU 1.80GHz × 8 processor,
    with 15.3 GiB of RAM, under Debian GNU/Linux 11.] In the log-scaled figure, we can observe the trend of computation
time as a function of the population size $N$, which is linear for the
randomized algorithm and quadratic for the full one (
 @alg-interaction). We can also see that the
randomized version of the algorithm is between 17 to 100 times faster
than the full one in this example, taking only 2 seconds in average for
the randomized version versus 211 seconds for 
@alg-interaction for the biggest population size
($N=2600$) and $T=500$.

![Full vs random algorithm computztion time](img/time_pop_size.png){#fig-fullvsrandom width=80%}

@@

![Example of Marked Poisson measure on $[0,T]$ with $m(\mathrm{d} t) = L \mathrm{d} t$ (jump times occur at Poisson arrival times of rate $L$) and with $\nu(\mathrm{d} y) = \frac{1}{L} \mathbf{1}_{[0, L]}(y) \mathrm{d} y$ (marks are drawn uniformly on $[0,L]$).
The mean measure is then $\mu(\mathrm{d} t, \mathrm{d} y) = \mathrm{d} t {1}_{[0,L]}(y) \mathrm{d} y$.](img/plot_poisson.tex){#fig-poisson}
